---
title: "Cody's Spotify Listening Habits - 2021"
output:
  github_document:
    html_preview: true
---

# Confessions of a Song Repeater

* A simple demonstration of how to analyze spotify streaming data, using my statistics from the last year.
* I have a pretty broad taste in music, but will often repeatedly listen to a new song or group when there's a particular sound that is striking. 

```{r include=FALSE}
## inspirations
# https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-02-25/readme.md
# https://towardsdatascience.com/explore-your-activity-on-spotify-with-r-and-spotifyr-how-to-analyze-and-visualize-your-stream-dee41cb63526
# https://datacarpentry.org/R-ecology-lesson/05-r-and-databases.html
# https://medium.com/@skyetetra/pretending-to-have-a-sql-database-with-rstudio-e80d9a1338b1
# https://cran.r-project.org/web/packages/dbparser/vignettes/Database_Saving.html
# https://www.youtube.com/watch?v=E2Ld3QdXYZo
# https://databricks.com/blog/category/engineering/open-source

##### create an in-memory sqlite database using DBI #####
# https://medium.com/@skyetetra/pretending-to-have-a-sql-database-with-rstudio-e80d9a1338b1

library(DBI)
library(jsonlite)
library(tidyverse)
library(ggplot2)
library(plotly)
library(DT)

con <- dbConnect(RSQLite::SQLite(), dbname = ":memory:")

path_data <- 'data/raw/Cody'

### list certain types of files desired
## this JSON is flat
stream_files <- list.files(path = path_data, 
                           pattern = 'StreamingHistory' , 
                           full.names=TRUE)

## this JSON is nested
playlist_files  <- list.files(path = path_data, 
                              pattern = 'Playlist' , 
                              full.names=TRUE)
```

## Transforming and Structuring Data

* The following function binds the streaming history files into one data.frame/table
   * The number of streaming files you have can vary, and depends on how many streams you have listened to in the last year.
   * A "stream" can be anything from listening to a song for 3 seconds (and then skipping it) to pressing the "play" button on an entire album

```{r}
##### convert flat JSON to CSV #####
fx_stackJSON <- function(file_list){
    
    out <- lapply(file_list, FUN = function(x){
        
        mid = fromJSON(x[[1]])
        
    }) %>% bind_rows(.)
    
    return(out)
    
}

# jsonlite::parse_json(<input_data>)
stream_data <- fx_stackJSON(stream_files)

## each row in this data frame has a list nested in the column "items"
playlist_data <- fromJSON(playlist_files, flatten=TRUE)

# names(stream_data)
# names(playlist_data)
```

* Below we use an anonymous function inside of `lapply()` to parse the playlist files, which are nested within one particular field as a list, and convert it into a nice flat data.frame/table

```{r}
## grab just the playlist names
playlist_names <- playlist_data$playlists$name

## extract the nested list as a data.frame, assign the playlist name
combined_playlists <- lapply(seq(1, length(playlist_data$playlists$items)), FUN = function(playlist_index){
    
    ## for each data.frame, add a column that matches the index of playlist_names
    playlist <- playlist_data$playlists$items[[playlist_index]]
    
    ## append the playlist name to the data.frame
    playlist$playlist_name <- playlist_names[playlist_index]
    
    ## cleanup column names
    playlist <- playlist %>% rename(trackName = track.trackName, 
                                    artistName = track.artistName,
                                    albumName = track.albumName, 
                                    trackUri=track.trackUri)
        
    return(playlist)
    
    ## return a single data.frame
}) %>% bind_rows(.)

##### load data to an in memory database #####
dbWriteTable(con, "StreamingHistory", stream_data)
dbWriteTable(con, "Playlists", combined_playlists)
```

```{r include=FALSE}
##### run queries with DBI #####
dbGetQuery(con, "SELECT * FROM StreamingHistory LIMIT 10")
dbGetQuery(con, "SELECT * FROM Playlists LIMIT 10")

##### run queries with dbplyr #####
# https://datacarpentry.org/R-ecology-lesson/05-r-and-databases.html

##### explore spotifyR #####
# https://towardsdatascience.com/explore-your-activity-on-spotify-with-r-and-spotifyr-how-to-analyze-and-visualize-your-stream-dee41cb63526
```

### Most Frequently Played Songs

```{r eval=FALSE}
## in SQL
dbGetQuery(con, "SELECT artistName, trackName, COUNT(trackName) as number
           FROM StreamingHistory
           GROUP BY trackName
           ORDER BY number DESC
           LIMIT 10")
```

```{r echo=FALSE}
##### most played songs - StreamingHistory ######

stream_data %>% 
    group_by(artistName, trackName) %>% 
    count %>% 
    arrange(desc(n)) %>% 
    head(10) %>% 
  datatable()
```

### Most Frequently Played Artists

```{r eval=FALSE}
## in SQL
dbGetQuery(con, 'SELECT artistName, COUNT(artistName) as count
            FROM StreamingHistory
            GROUP BY artistName 
            ORDER BY count DESC
            LIMIT 10')
```

```{r echo=FALSE}
##### most played artists #####
res_freq_artists <- stream_data %>% 
    group_by(artistName) %>% 
    count %>% 
    arrange(desc(n)) %>% 
    head(10) 
res_freq_artists %>% 
    datatable()
```

### Most Songs Played in One Day   

* This would include skipping a song after, say, 3-10 seconds 
    * Could filter out results that are less than 30 seconds to 1 minute

```{r echo=FALSE}
##### most songs in one day #####
stream_data %>% mutate(date = str_sub(endTime, 1, 10)) %>% 
    group_by(date) %>% 
    count %>% 
    arrange(desc(n)) %>% 
    head(10) %>% 
    datatable()
```

```{r eval=FALSE}
## this is SQLITE syntax, other SQL might use "LEFT(endTime, 10)"
# temporarily create a column named date, use that to group by...COUNT doesn't work on aliases I guess
dbGetQuery(con, 'SELECT substr(endTime, 1, 10) AS date, COUNT(endTime) AS count, SUM(msPlayed) as songTime
                 FROM StreamingHistory
                 GROUP BY date
                 ORDER BY count DESC
                 LIMIT 10')
```

### Of top 10 listened to artists, over what time period did I listen to them the most? 

* Counts of that artist per endDate through time
* This is much nicer to view as an interactive ggplotly chart (but GitHub doesn't quite render it)
* `NOTE` Opening this file in the `.html` format will display the following chart interactively. For instance, you can double-click a band name in the legend to spotlight those particular data points. 

```{r warning=FALSE}
plotly::ggplotly(
    stream_data %>% mutate(date = str_sub(endTime, 1, 10)) %>% 
    # this is a shortcut subquery, technically
    # filter out song listens that are less than 1 minute in length
    filter(artistName %in% res_freq_artists$artistName & msPlayed > (60 * 1000)) %>% 
    group_by(date, artistName) %>% 
    count %>% 
    ggplot(aes(x = as.Date(date), y = n, group = artistName, color = artistName)) +
    geom_line() + 
    geom_point() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    scale_x_date(date_breaks = "months" , date_labels = "%b-%y") +
    xlab('Month-Year') + 
    ylab('Number of Song Listens (> 60 seconds)') + 
    ggtitle('Top 10 Artists of the Year: Number of Listens Per Day')
)

```

```{r include=FALSE}
# an alternative method of using dplyr to join the data together, like a SQL subquery
# 
# ## subquery example, joining IS a form of filtering by another table
# subquery <- stream_data %>% 
#     mutate(date = str_sub(endTime, 1, 10)) %>% 
#     # this is a shortcut subquery, technically
#     # filter out song listens that are less than 1 minute in length
#     right_join(., stream_data %>%
#         filter(msPlayed > (60 * 1000)) %>% 
#         group_by(artistName) %>% 
#         count %>% 
#         arrange(desc(n)) %>% 
#             select(artistName) %>% 
#         head(10))
```


```{r include=FALSE}
### Seasonal Song Repetition

# Kowloon, MadeinTYO, Modern Baseball, No vacation, Sea Power

    stream_data %>% mutate(date = str_sub(endTime, 1, 10)) %>% 
    # this is a shortcut subquery, technically
    # filter out song listens that are less than 1 minute in length
    filter(artistName %in% res_freq_artists$artistName & msPlayed > (60 * 1000)) %>% 
    group_by(date, artistName) %>% 
    count %>% 
    ggplot(aes(x = as.Date(date), y = n, group = artistName, color = artistName)) +
    geom_line() + 
    geom_point() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    scale_x_date(date_breaks = "months" , date_labels = "%b-%y") +
    xlab('Month-Year') + 
    ylab('Number of Song Listens (> 60 seconds)') + 
    ggtitle('Top 10 Artists of the Year: Number of Listens Per Day')

```

# Changes in Listening Behavior, Summer 2019 vs. 2021

* Spotify only provides listening data for the past year, but if you have ever saved your "Summer Rewind 20XX" or other auto-generated playlists then you may use those to see if your tastes have changed through time.
* My "Summer Rewind 2019" playlist in particular is useful to analyze from a song repeater perspective
* Let's look at this in two ways:
  * Which artists in 2019 am I still listening to (or not)?
  * Of the top 50 artists I listened to in 2021, did I listen to them in 2021?

```{r  warning=FALSE}
## grab the Summer 2019 playlist
previous_repeat_plays <- combined_playlists %>% filter(playlist_name == 'Summer Rewind 2019')

## grab total listens per artist THIS year, then join to compare
current_repeat_plays <- stream_data %>% 
    group_by(artistName, trackName) %>% 
    count %>% 
    arrange(desc(n)) 

```

  * Which artists in 2019 am I still listening to (or not)?

```{r}
datatable(data.frame(Artist2019 = previous_repeat_plays$artistName,
                     Listened_to_2021 = previous_repeat_plays$artistName %in% current_repeat_plays$artistName))
```

* Apparently I'm still listening to all 50 artists I enjoyed Summer of 2019.
* Are any of the 2019 artists _still_ in my Top 50 for 2021 though?

```{r warning=FALSE}
# run an inner join
current_repeat_plays %>% head(50) %>% inner_join(., previous_repeat_plays, by = 'artistName') %>% datatable()
```

* Only one frequently played artist in 2019 is still in my top 50 this year. 

```{r include=FALSE}
##### compare most played artists this year to Summer Rewind playlists of 2019 #####

##### artist diversity ######

##### song diversity #####
```